Automated iterative forecasting for the Portal Project

### 2018-01-23 manuscript/portal_iterative_forecasting_ms.Rmd

Implementation of iterative forecasting requires the regular rebuilding of models with new raw data as it becomes available and the presentation of those forecasts in usable forms; in our case, this occurs monthly. Doing this in an efficient and maintainable way relies on developing an automated pipeline to handle the six stages of converting raw data into new forecasts: data collection, data sharing, data manipulation, modeling and forecasting, archiving, and presentation of the forecasts (Figure 1). We implemented this pipeline using a "continuous analysis" framework [*sensu* @beaulieu2017] that automatically processes the most up-to-date data, refits the models, makes new forecasts, archives the forecasts, and updates a website with analysis of current and previous forecasts. In this section we describe our approach to streamlining and automating the multiple components of the forecasting pipeline.

![Figure 1. Forecasting pipeline](figs/pipeline_fig.png)

## Continuous Analysis Framewor

A core component of iterative near-term forecasting is the regular rerunning of forecasting pipeline. This can be conducted manually - with a human making sure all code is run and all tables and files are updated - or automatically by having the computer conduct those tasks. We chose to have the computer run our pipeline and employed "continuous analysis" [*sensu* @beaulieu2017] to drive the automation of both the full pipeline and a number of its individual components. Continuous analysis uses a set of tools originally designed for software development called "continuous integration" (CI). CI combines computing environments for running code with monitoring systems to identify changes in data or code. Essentially, CI is a computer helper whose job is to watch the pipeline and, when it sees a change in the code or data, it runs all the tasks needed to ensure that the forecasting pipeline runs from beginning to end. This is useful for iterative near-term forecasting because it does not rely on humans to remember to create forecasts when new models or data are added. These tools are common in the area of software development where they are used to automate software testing and integrate work by multiple software developers working on the same software. However, these tools can be used for any computational task that needs to be regularly repeated or run after changes to code or data [@beaulieu2017]. Because of the widespread use of CI in software development, several CI services already exist. Our forecasting pipeline currently runs on a publicly available continuous integration service (Travis CI; https://travis-ci.org/) that is free for open source projects (up to a limited amount of computing time); alternative services that can be used to run code on local or cloud-based computational infrastructure are available, such as Drone (http://try.drone.io/) [@beaulieu2017]. As detailed below, we use CI to quality check data, test code using "unit tests" [@wilson2014], build models, make forecasts, and publicly present and archive the results (Figure 2).

![Figure 2. Continuous integration system.](figs/contanaly_fig.png)

@@

Iterative forecasting benefits from frequently updated data on the state of the system so that state changes can be quickly incorporated into new forecasts [@dietze2018]. Frequent data collection and rapid entry and processing of data are both important for providing the most up-to-date data for forecasting. Since our data is collected monthly, ensuring that the models have access to the newest data requires a data latency period of less than 1 month from collection to availability for modeling. To accomplish this, we automated components of the data processing and quality assurance/quality control (QA/QC) process to reduce the time needed to add new data to the database.

@@

The Portal Project has a long history of making its data publicly available, which means that anyone can use it for forecasting or other projects. Historically the publication of the data was conducted through data papers, the most common approach in ecology; however, this approach caused years of data latency. Recently, the project has switched to posting data directly to a public GitHub repository with a CC0 license. This immediate posting reduces that data latency to less than one month and, therefore, makes meaningful iterative near-term forecasting possible for not only our group but other interested parties, as well.

@@

Once data is available, it needs to be processed into a form appropriate for modeling. In many ecological datasets, this requires not only simple data manipulation but also a good understanding of the dataset to allow data to be aggregated appropriately. These data manipulation steps are often conducted using custom one-off code to convert the data into the desired form [@morris2013], but this approach has several limitations. First, each researcher must develop their own data manipulation code, which is inefficient and can result in different decisions by different researchers about the details of data cleaning and aggregation. Subtle differences in data processing decisions have lead to confusion when reproducing results for the Portal data in the past. Second, this kind of code is rarely robust to changes in data structure and location. Based on our experience developing and maintaining the Data Retriever [@morris2013; @senyondo2017], these kinds of changes are common. Finally, this kind of code is generally poorly tested, which can lead to errors based on mistakes in data manipulation. To avoid these issues for the Portal Project data, the Portal team has been developing an R package (portalR; http://github.com/weecology/portalr) for acquiring the data and handling common data cleaning and aggregation tasks. As a result, our modeling and forecasting code only needs to install this package and run the data manipulation and summary functions to get the appropriate data. The package undergoes thorough automated unit tests to ensure that data manipulations are achieving the desired results. Having data manipulation code maintained in a separate package that focuses on consistently providing properly summarized forms of the most recent data has made maintaining the forecasting code itself much more straightforward.

@@

Ideally, iterative near-term forecasting involves regularly refitting a variety of different models. New models should be easy to incorporate to allow for iterative improvements to the general modeling structure and approach. We use CI to refit the models and make new forecasts each time the modeling code changes and when new data becomes available. We use a plugin infrastructure to allow new models to be easily added to the system. Each model is a script that takes in standardized inputs and returns standardized outputs (Figure 3). All models are run by the main forecasting code at each update, and the standardized outputs are combined to store the results of the different models' forecasts. A weighted ensemble model is then added with weights based on how well individual models fit the training data. This plugin infrastructure allows flexibility in all aspects of the modeling process, making it easier to explore new modeling approaches, and allows new models that fit the data well to immediately improve the ensemble forecast.

![Figure 3. Demonstration of plugin infrastructure where each model script uses data provided by the core forecasting code and returns its forecasts in a predefined structure.](figs/plugin_fig.gif)

@@

## Evaluation and Hindcasting

@@

Publicly archiving forecasts before new data is collected allows the field to assess, compare, and build on forecasts made by different groups [@mcgill2012; @tredennick2016; @dietze2017; @harris2018]. This archiving serves as a form of pre-registration for model predictions, helping facilitate unbiased interpretation of model performance. We explored three major repositories for archiving our forecasts: FigShare, Zenodo, and Open Science Framework. While all three repositories allowed for easy manual submissions (i.e., a human uploading files after each forecast), automating this process was substantially more difficult. Various combinations of repositories, APIs and associated R packages had issues with: 1) integrating authorization with continuous integration; 2) automatically making archived files public; 3) adding new files to an existing location; or 4) automatically permanently archiving the files. Our eventual solution was to leverage the GitHub-Zenodo integration (https://guides.github.com/activities/citable-code/) and automatically push forecasts to the GitHub repository from the CI server. The GitHub-Zenodo integration is designed to automatically create versioned archives of GitHub repositories. There is an existing one-time process for linking our forecasting repository on  GitHub with Zenodo. Once this link is created, each time a new forecast is created, our pipeline adds the new forecasts to the GitHub repository and uses the GitHub API to create a new "release" of our repository. This triggers the Zenodo-GitHub integration, which automatically archives the resulting forecasts and the code that generated them under a top-level DOI that refers to all archived forecasts ([https://doi.org/10.5281/zenodo.833438](https://doi.org/10.5281/zenodo.833438)). Through this process, we automatically archive every forecast made with a documented history of the archive. While this approach is functional because everything in the repository is archived when a new forecast is made, these archives are complicated, making it more complicated than necessary to find and access the forecasting results.

@@

![Figure 4. Screen capture of the homepage of the Portal Forecasting website; http://portal.naturecast.org](figs/website_fig.png)

@@

Following the recommendations of Dietze et al [-dietze2018], we developed an automated iterative forecasting system to support repeated forecasting of an ecological system. Our forecasting system automatically acquires and processes the newest data, refits the models, makes new forecasts, publicly archives those forecasts, and presents both the current forecast and information on how previous forecasts performed. Every week our forecasting system generates a new set of forecasts with no human intervention, except for the entry of new field data. This ensures that forecasts based on the most recent data are always available and allows us to rapidly assess the performance of multiple forecasting models for a number of different states of the system, including the abundances of individual species and community-level variables such as total abundance. To create this iterative near-term forecasting system, we used R to process data and conduct analyses, and we leveraged already existing services (i.e. GitHub, Travis, Docker) for more complicated cyberinfrastructure tasks. Thus, our approach to developing iterative near-term forecasting infrastructure provides an example for how short-term ecological forecasting systems can be initially developed.

>>>>

Implementation of iterative forecasting requires the regular rebuilding of models with new raw data as it becomes available and the presentation of those forecasts in usable forms; in our case, this occurs monthly. Doing this in an efficient and maintainable way relies on developing an automated pipeline to handle the six stages of converting raw data into new forecasts: data collection, data sharing, data manipulation, modeling and forecasting, archiving, and presentation of the forecasts (Figure 1). To implement the pipeline outlined in Figure 1, we used a "continuous analysis" framework [*sensu* @beaulieu2017] that automatically processes the most up-to-date data, refits the models, makes new forecasts, archives the forecasts, and updates a website with analysis of current and previous forecasts. In this section we describe our approach to streamlining and automating the multiple components of the forecasting pipeline and the tools and infrastructure we employed to execute each stage of the pipeline.

![Figure 1. Stages of the forecasting pipeline. To go from raw data to forecast presentation involves a number of stages, each of which required unique tasks, tools and infrastructure. While each stage has unique tasks associated with it, the stages of our pipeline are also interdependent. Our pipeline is primarily a linear structure where outputs from one stage form the inputs for the subsequent stage ](figs/pipeline_fig.png)

## Continuous Analysis Framework

A core component of iterative near-term forecasting is the regular rerunning of the forecasting pipeline. This can be conducted manually---with a human making sure all code is run and all tables and files are updated---or automatically by having the computer conduct those tasks. We chose to have the computer run our pipeline and employed "continuous analysis" [*sensu* @beaulieu2017] to drive the automation of both the full pipeline and a number of its individual components. Continuous analysis uses a set of tools originally designed for software development called "continuous integration" (CI). CI combines computing environments for running code with monitoring systems to identify changes in data or code. Essentially, CI is a computer helper whose job is to watch the pipeline and, when it sees a change in the code or data, it runs all the tasks needed to ensure that the forecasting pipeline runs from beginning to end. This is useful for iterative near-term forecasting because it does not rely on humans to remember to create forecasts when new models or data are added. These tools are common in the area of software development where they are used to automate software testing and integrate work by multiple software developers working on the same software. However, these tools can be used for any computational task that needs to be regularly repeated or run after changes to code or data [@beaulieu2017]. Because of the widespread use of CI in software development, several CI services already exist. Our forecasting pipeline currently runs on a publicly available continuous integration service (Travis CI; https://travis-ci.org/) that is free for open source projects (up to a limited amount of computing time); alternative services that can be used to run code on local or cloud-based computational infrastructure are available, such as Drone (http://try.drone.io/) [@beaulieu2017]. As detailed below, we use CI to quality check data, test code using "unit tests" [@wilson2014], build models, make forecasts, and publicly present and archive the results (Figure 2).

![Figure 2. Continuous integration system. Each box denotes the core infrastructure used for each stage of the forecasting pipeline denoted in Figure 1. Continuous integration, denoted here with the Travis icon (a girl wearing safety glasses and hardhat), plugs into our pipeline at every stage.  Travis triggers the code involved in major events that involve integration across stages of the pipeline, such as taking the output from the forecasting stage (purple box) to create an updated presentation (rose box). Travis also runs tasks within a stage, such as conducting tests to make sure code changes have not introduced errors (icons on arrows originating and ending on the same box)](figs/contanaly_fig.png)

@@

Iterative forecasting benefits from frequently updated data on the state of the system so that state changes can be quickly incorporated into new forecasts [@dietze2018]. Frequent data collection and rapid entry and processing of data are both important for providing the most up-to-date data for forecasting. Since our data is collected monthly, ensuring that the models have access to the newest data requires a data latency period of less than 1 month from collection to availability for modeling. To accomplish this, we automated components of the data processing and quality assurance/quality control (QA/QC) process to reduce the time needed to add new data to the database (Figures 1 and 2).

@@

The Portal Project has a long history of making its data publicly available, which means that anyone can use it for forecasting or other projects. Historically the publication of the data was conducted through data papers, the most common approach in ecology; however, this approach caused years of data latency. Recently, the project has switched to posting data directly to a public GitHub repository with a CC0 license (Figure 1). This immediate posting reduces that data latency to less than one month and, therefore, makes meaningful iterative near-term forecasting possible for not only our group but other interested parties as well.

@@

Once data is available, it needs to be processed into a form appropriate for modeling (Figure 1). In many ecological datasets, this requires not only simple data manipulation but also a good understanding of the dataset to allow data to be aggregated appropriately. These data manipulation steps are often conducted using custom one-off code to convert the data into the desired form [@morris2013], but this approach has several limitations. First, each researcher must develop their own data manipulation code, which is inefficient and can result in different decisions by different researchers about the details of data cleaning and aggregation. Subtle differences in data processing decisions have lead to confusion when reproducing results for the Portal data in the past. Second, this kind of code is rarely robust to changes in data structure and location. Based on our experience developing and maintaining the Data Retriever [@morris2013; @senyondo2017], these kinds of changes are common. Finally, this kind of code is generally poorly tested, which can lead to errors based on mistakes in data manipulation. To avoid these issues for the Portal Project data, the Portal team has been developing an R package (portalR; http://github.com/weecology/portalr) for acquiring the data and handling common data cleaning and aggregation tasks. As a result, our modeling and forecasting code only needs to install this package and run the data manipulation and summary functions to get the appropriate data (Figure 2). The package undergoes thorough automated unit tests to ensure that data manipulations are achieving the desired results. Having data manipulation code maintained in a separate package that focuses on consistently providing properly summarized forms of the most recent data has made maintaining the forecasting code itself much more straightforward.

@@

Ideally, iterative near-term forecasting involves regularly refitting a variety of different models (Figure 1). New models should be easy to incorporate to allow for iterative improvements to the general modeling structure and approach. We use CI to refit the models and make new forecasts each time the modeling code changes and when new data becomes available (Figure 2). We use a plugin infrastructure to allow new models to be easily added to the system. Each model is a script that takes in standardized inputs and returns standardized outputs (Figure 3). All models are run by the main forecasting code at each update, and the standardized outputs are combined to store the results of the different models' forecasts. A weighted ensemble model is then added with weights based on how well individual models fit the training data. This plugin infrastructure allows flexibility in all aspects of the modeling process, making it easier to explore new modeling approaches, and allows new models that fit the data well to immediately improve the ensemble forecast.

![Figure 3. Demonstration of plugin infrastructure where each model script (represented here as AR1.R) uses data provided by the core forecasting code (represented here as rodent.csv) and returns its forecasts in a predefined structure that is consistent across models (represented here as 2017_12_08forecasts.csv) to allow cross-model evaluations.](figs/plugin_fig.gif)

@@

@@

Publicly archiving forecasts before new data is collected allows the field to assess, compare, and build on forecasts made by different groups [@mcgill2012; @tredennick2016; @dietze2017; @harris2018](Figure 1). This archiving serves as a form of pre-registration for model predictions, helping facilitate unbiased interpretation of model performance. We explored three major repositories for archiving our forecasts: FigShare, Zenodo, and Open Science Framework. While all three repositories allowed for easy manual submissions (i.e., a human uploading files after each forecast), automating this process was substantially more difficult. Various combinations of repositories, APIs and associated R packages had issues with: 1) integrating authorization with continuous integration; 2) automatically making archived files public; 3) adding new files to an existing location; or 4) automatically permanently archiving the files. Our eventual solution was to leverage the GitHub-Zenodo integration (https://guides.github.com/activities/citable-code/) and automatically push forecasts to the GitHub repository from the CI server. The GitHub-Zenodo integration is designed to automatically create versioned archives of GitHub repositories. There is an existing one-time process for linking our forecasting repository on  GitHub with Zenodo. Once this link is created, each time a new forecast is created, our pipeline adds the new forecasts to the GitHub repository and uses the GitHub API to create a new "release" of our repository. This triggers the Zenodo-GitHub integration, which automatically archives the resulting forecasts and the code that generated them under a top-level DOI that refers to all archived forecasts ([https://doi.org/10.5281/zenodo.833438](https://doi.org/10.5281/zenodo.833438)). Through this process, we automatically archive every forecast made with a documented history of the archive. While this approach is functional because everything in the repository is archived when a new forecast is made, these archives are complicated, making it more complicated than necessary to find and access the forecasting results.

@@

![Figure 4. Screen capture of the homepage of the Portal Forecasting website; http://portal.naturecast.org. This site contains information on the most current forecasts, evaluation of forecast performance, and general information about the species being forecast.](figs/website_fig.png)

@@

Following the recommendations of Dietze et al [-dietze2018], we developed an automated iterative forecasting system (Figures 1 and 2) to support repeated forecasting of an ecological system. Our forecasting system automatically acquires and processes the newest data, refits the models, makes new forecasts, publicly archives those forecasts, and presents both the current forecast and information on how previous forecasts performed. Every week our forecasting system generates a new set of forecasts with no human intervention, except for the entry of new field data. This ensures that forecasts based on the most recent data are always available and allows us to rapidly assess the performance of multiple forecasting models for a number of different states of the system, including the abundances of individual species and community-level variables such as total abundance. To create this iterative near-term forecasting system, we used R to process data and conduct analyses, and we leveraged already existing services (i.e. GitHub, Travis, Docker) for more complicated cyberinfrastructure tasks. Thus, our approach to developing iterative near-term forecasting infrastructure provides an example for how short-term ecological forecasting systems can be initially developed.
